Combinatorial interpretations of~holonomic recurrences

### 2021-09-13 article/main.tex
Fix brutal approach

\title{Combinatorial interpretations of holonomic recurrences}
\author{Bartłomiej Puget}

@@

\DeclareMathOperator{\textiff}{\text{iff}}
\DeclareMathOperator{\N}{\mathbb{N}}
\DeclareMathOperator{\poly}{\mathbb{P}}
\DeclareMathOperator{\troot}{\text{root}}
\DeclareMathOperator{\tleft}{\text{left}}
\DeclareMathOperator{\tright}{\text{right}}
\DeclareMathOperator{\tLeft}{\text{Left}}
\DeclareMathOperator{\tRight}{\text{Right}}
\DeclareMathOperator{\tSub}{\text{Sub}}
\DeclareMathOperator{\tPre}{\text{Pre}}
\DeclareMathOperator{\tSuf}{\text{Suf}}
\DeclareMathOperator{\n}{\bullet}
\DeclareMathOperator{\no}{\o}

@@

Generating functions is a way of describing infinite sequences using formal power series, where coefficients are encoding the the consecutive elements. It allows us to incorporate many advanced methods of analysis of encoded objects.

Every recursive structure can be described using a~generating function equation, which can be then solved and transformed into a~holonomic recurrence, i.e.~recurrence that in order to define \(n\)-th element of the sequence requires just finite number of previous ones and coefficients that are polynomials of \(n\).

It allows us to quickly count objects of given structure with given size. At first glance, the recurrence seems to be closely related to the structure itself. It just seems like it should describe simple transformations resulting in constructing bigger objects from slightly smaller ones. In case of trees it could be for example adding vertices. However, there is no known method of interpreting it in the language of the combinatorial structures.

@@

The initial inspiration for the thesis came from the publication \textit{A natural counting of lambda terms}~\cite{inspiration}, which shows bijections among seemingly different tree structures, which are lambda trees, black-white trees~\cite{blackwhite} and zigzag free trees~\cite{inspiration}\todo{Find better cite}. They share the same holonomic recurrence, but there is no known immediate connection between the holonomic recurrence and the structures itself.

Lambda terms are very popular structures used for analyzing functions and programming languages. Their very simple recursive definition makes them very attractive. Better understanding them could allow us improving programming language compilers and in consequence improve every aspect of our lives. The interpretation of holonomic recurrence for lambda trees itself could allow us to discover new ways of generating random lambda terms via extending smaller ones in similar way to Rémy's algorithm for sampling binary trees~\cite{remy,note}.

@@

Section~\ref{sec:brutal_iterative_approach} presents our final approach to the original problem, i.e.~computer-aided iterative approach that allowed us to find the intermediate solution. We believe that it is possible to extend this idea in order to obtain the full solution, but we were not able to achieve that due to the limited time that was available to us.

@@

\(\troot(T)\) is implicitly defined as a~root of tree \(T\).

\(\tleft(v)\) is implicitly defined as a~left child of node \(v\).

\(\tright(v)\) is implicitly defined as a~right child of node \(v\).

\(\tSub(v)\) is implicitly defined as a~whole subtree rooted with node \(v\).

\(\tLeft(v)\) is implicitly defined as a~left subtree of node \(v\).

\(\tRight(v)\) is implicitly defined as a~right subtree of node \(v\).

Sibling of node \(v\) is defined as node \(u\) having the same parent as \(v\).

Uncle of node \(v\) is defined as a~sibling of the parent of \(v\).

@@

We say that node \(v\) is twinling \(\textiff\) \(v\) has at least one sibling.

@@

\(x_i\) is implicitly assumed to be an~\(i\)-th element of the sequence \(X\), where \(x\) can be any lower-case letter and \(X\) is upper-case \(x\).

@@

\begin{definition}
    The ordinary generating function of sequence \(A = {(a_i)}_{i=0}^{\infty}\) is defined as a~formal power series:
\[\gf{A}(z) = \sum_{i=0}^{\infty} a_i z^i.\]
\end{definition}

Note that in order to compute the \(k\)-th element of the sequence, namely \(a_k\), we just need to find a~constant corresponding to indeterminate \(z^k\).

@@

\begin{example}%
    \label{ex:bin_gf}
    Let's consider recursive definition of binary trees shown in Figure~\ref{fig:binary_recursion}. We can create a~functional equation to count trees with exactly n internal nodes. To do so, we make internal nodes contribute to the generating function with \(z^1 = z\) and leaves with \(z^0 = 1\):

@@

In the thesis, if not stated otherwise, we will be working on rooted ordered unlabeled \(n\)-ary trees. To understand the terms, let's define them:
\begin{itemize}
    \item rooted tree is a~tree with single vertex marked as root. This allows us to define hierarchy of nodes using such terms as parent and child;
    \item ordered trees introduce order on children from the leftmost to the rightmost;
    \item unlabeled trees have nodes that are not distinguishable by their label. We will introduce soon kinds of nodes, but you shouldn't treat them as labels, but rather way to describe the structure of the node that is already present (e.g.~we can call node with just left child an~\(l\) node, but it is not its label. If the right child was added to the \(l\) node, it would not be an~\(l\) node anymore).
\end{itemize}
Additionally we allow that \(n\)-ary node has some of its children unset (e.g.~we can define a~binary node with just left child set, but not right one).

@@

\begin{definition}
    For single vertex \(v\), the pre-order notation is defined as \(\enc{v}\).
\end{definition}

@@

\begin{definition}
    Let \(n\) be the length of the sequence \(\enc{T}\). Notation of suffix of \(T\) in relation to \(v\), namely \(\enc{\tSuf(T, v)}\), is defined as a~suffix of \enc{T} of size \(n - \size{\enc{\tPre(T, v)}} - \size{\enc{\tSub(v)}}\).
\end{definition}

@@

These interpretations have been source of ideas that that we were incorporating in order to make our partial interpretation for the case of lambda trees in unary de Bruijn notation.

@@

Most of the time we are operating on trees with some nodes of weight \(0\) and some of weight \(1\). We may as well try to make them variable or assign some other values (e.g.~small prime numbers).

@@

To show the idea of such transformations, let's define transformation context with holes. The context specifies shape of the fragment of the tree. It can have holes that specify whole subtree. We can also specify whether the root of the context has parent and is it the left or right child of its parent.

\begin{example}
    We want to modify a node being right a child of its parent. We do not care what is below that node (i.e.~if that is a leaf or if it has children). We want to add a new node of which the original one is left child. We do not want the new node to have a right child. Additionally we want to swap potential children of the original node. We can define following transformation:
    \begin{center}
        \includeinlinescaledsvg{.4}{.7}{lambda__contexts__def_001}%
        \(\Rightarrow\)
        \includeinlinescaledsvg{.4}{.7}{lambda__contexts__def_002}%
    \end{center}
\end{example}

We marked the node being transformed to clarify the picture.

@@

\begin{itemize}
    \item node is a~virtual leaf (each tree with \(n\) actual nodes has exactly \((n + 1)\) places where such contex apply):\\
        \includeinlinescaledsvg{1}{.7}{lambda__contexts__type_001}%
    \item node is an~actual node (tree with \(n\) actual nodes has excatly \(n\) places where such context apply):\\
        \includeinlinescaledsvg{1}{.7}{lambda__contexts__type_002}%
    \item node is a~root (each tree has exactly \(1\) place where such context apply):\\
        \includeinlinescaledsvg{1}{.7}{lambda__contexts__type_003}%
    \item node has its parent on the left, i.e.~is right child of its parent:\\
        \includeinlinescaledsvg{1}{.7}{lambda__contexts__type_004}%
    \item node has its parent on the right, i.e.~is the left child of its parent:\\
        \includeinlinescaledsvg{1}{.7}{lambda__contexts__type_005}%
\end{itemize}

@@

We will be operating on transformation classes, sections and parts. Let's define them:

\begin{definition}
    Class \(T_i\) is defined as a~set of transformations of trees of size \((n - i)\) into trees of size \(n\).
\end{definition}

\begin{definition}
    Section \(T_{i,j}\) is defined as a~subset of \(T_i\) adding a~specific set of nodes. We assign \(j\) according to the lexicographical order of lexicographically sorted node types.
\end{definition}

\begin{definition}
    Part \(T_{i,j,k}\) is defined as a~single transformation in section \(T_{i,j}\). \(k\) is assigned according to the order of discovering the transformation.
\end{definition}

In order to make interpretation more human readable, we will be using local notation:

\begin{definition}
    In the context of \(T_i\), \(k\) is defined as \((n - i)\).
\end{definition}

\begin{definition}
    In the context of \(T_i\), \(C_k\) is defined as \(T(n - i)\), i.e.~number of tree transformations from \(T_i\) are working on.
\end{definition}

\begin{definition}
    Section equation \(E_{i, j}\) is an~equation defined in the context of \(T_i\) that evaluates to a~total number of trees that should be generated by transformations from \(T_{i, j}\).
\end{definition}

\begin{definition}
    Section expression \(S_{i, j}\) is a~pair of section equations and a~set of kinds of nodes the transformation introduces.
\end{definition}

For the problem of interpreting trees of lambda terms we define following section expressions:

\[\begin{array}{rccrlc}
        S_{0,0} &=& (& (k + 1) C_k,& \emptyset &)\\
        S_{1,0} &=& (& - (2 k + 1) C_k,& \{l\} &)\\
        S_{1,1} &=& (& - 2 (k + 1) C_k,& \{s\} &)\\
        S_{2,0} &=& (& - (4 k) C_k,& \{ao\} &)\\
        S_{2,1} &=& (& (k) C_k,& \{ll\} &)\\
        S_{2,2} &=& (& (4 k + 2) C_k,& \{ls\} &)\\
        S_{2,3} &=& (& (k + 1) C_k,& \{ss\} &)\\
        S_{3,0} &=& (& 2 (2 k + 1) C_k,& \{aos\} &)\\
        S_{3,1} &=& (& - (2 k) C_k,& \{lls\} &)\\
        S_{3,2} &=& (& - (2 k + 1) C_k,& \{lss\} &)\\
        S_{4,0} &=& (& k C_k,& \{llss\} &)\\
\end{array}\]

Using context-based variables makes the equation simpler to interpret, as we can immediately see the relation between \(k\) and multiplier.

@@

As we want to make our work kind of organized, we are going to divide the process of finding interpretation into iterations. Each iteration should be a~small step into the direction of solving the original problem.

We are going to generate two kinds of trees: positive and negative ones. When computing the total number of trees of a given size, we count positive trees as \(1\) and negative as \(-1\). Our goal is to generate \(E_{i, j}\) trees for each section \(T_{i, j}\).

Our brains like small examples, so we will consider small \(n = 3\). This will cause some classes to be empty at the beginning, but it's ok. That's less trees to look at.

\begin{lstlisting}
=== Classes stats ===
[ ]  0,0:       0 of      16 | (k + 1) * Ck              []
[ ]  1,0:       0 of     -10 | - (2 * k + 1) * Ck        [l]
[ ]  1,1:       0 of     -12 | - 2 * (k + 1) * Ck        [s]
[ ]  2,0:       0 of      -4 | - (4 * k) * Ck            [ao]
[ ]  2,1:       0 of       1 | (k) * Ck                  [ll]
[ ]  2,2:       0 of       6 | (4 * k + 2) * Ck          [ls]
[ ]  2,3:       0 of       2 | (k + 1) * Ck              [ss]
[ ]  3,0:       0 of       2 | 2 * (2 * k + 1) * Ck      [aos]
[x]  3,1:       0 of       0 | - (2 * k) * Ck            [lls]
[ ]  3,2:       0 of      -1 | - (2 * k + 1) * Ck        [lss]
[x]  4,0:       0 of       0 | k * Ck                    [llss]
\end{lstlisting}

This statistic tells us for each section, how many trees have been generated of the expected total number of trees in that section. \verb|[x]| means that we generated enough trees for such a~section.

We start with interpreting \(E_{0,0} = (k + 1) C_k\). From statistics above, we see that we should generate \(16\) trees. \((k + 1)\) suggests pointing to a~virtual leaf. Let's do so.

@@

\begin{lstlisting}
# define 0,0: (k + 1) * Ck []
=== Classes stats ===
[x]  0,0:      16 of      16 | (k + 1) * Ck              []
[ ]  1,0:       0 of     -10 | - (2 * k + 1) * Ck        [l]
[ ]  1,1:       0 of     -12 | - 2 * (k + 1) * Ck        [s]
[ ]  2,0:       0 of      -4 | - (4 * k) * Ck            [ao]
[ ]  2,1:       0 of       1 | (k) * Ck                  [ll]
[ ]  2,2:       0 of       6 | (4 * k + 2) * Ck          [ls]
[ ]  2,3:       0 of       2 | (k + 1) * Ck              [ss]
[ ]  3,0:       0 of       2 | 2 * (2 * k + 1) * Ck      [aos]
[x]  3,1:       0 of       0 | - (2 * k) * Ck            [lls]
[ ]  3,2:       0 of      -1 | - (2 * k + 1) * Ck        [lss]
[x]  4,0:       0 of       0 | k * Ck                    [llss]
=== Diff stats ===
0,0:     0    16
\end{lstlisting}

The framework informs us that we have generated all trees for section \(0,0\) for \(n = 3\). It also can provide us with the visualization of the trees we have generated. For example:

\includeinlinesvg{.19}{lambda__trees_00__4_base}%
\(\Rightarrow\)
\includeinlinesvg{.19}{lambda__trees_00__4}%
\includeinlinesvg{.19}{lambda__trees_00__5}%
\includeinlinesvg{.19}{lambda__trees_00__6}%
\includeinlinesvg{.19}{lambda__trees_00__7}%

If we take a~look at more examples from class \(0\), we can start seeing some patterns:

@@

The order may be a~bit suggestive. Section \(1,0\) adds a~single \(l\) node, whereas section \(1,1\) adds a~single \(s\) node. It actually suggests how we can try to generate the first and second line of the trees above. The first one is simple. Let's add \(l\) node with pointed right virtual leaf above each actual node:

\begin{lstlisting}
c.define_class(1)

c.append_r_class_subtree_visitor(-1,                # negative node
    lambda kind, left, right:
        None if kind[0] is None else                # we ignore all virtual nodes
        ["l", kind] + left + right + [(None, "*")]  # for each actual node we add
                                                    # l node with right virtual
                                                    # node pointed as a parent
)
\end{lstlisting}
\todo{Repeat in natural language}

It creates following trees:
\todo{Mark transformed nodes}

\includeinlinesvg{.24}{lambda__trees_100__0_base}\(\Rightarrow\)\includeinlinesvg{.24}{lambda__trees_100__0}%
\includeinlinesvg{.24}{lambda__trees_100__1_base}\(\Rightarrow\)\includeinlinesvg{.24}{lambda__trees_100__1}%

\includeinlinesvg{.24}{lambda__trees_100__2_base}\(\Rightarrow\)\includeinlinesvg{.24}{lambda__trees_100__2}%
\includeinlinesvg{.24}{lambda__trees_100__3_base}\(\Rightarrow\)\includeinlinesvg{.24}{lambda__trees_100__3}%

It addresses all cases from the first line, but also introduces an~incorrect lambda tree in which \(s\) is the parent of \(l\), which cannot happen. We will fix it with the section \(2,2\) generating positive trees, which introduces one \(l\) node and one \(s\) by extending the tree with \(2\) nodes where \(s\) is parent of \(l\).

Next, we can take care of next cases of section \(0, 0\). It can be also expressed using sections \(1, 0\) and \(1, 1\):

\begin{lstlisting}
c.define_class(1)

# old visitor
c.append_r_class_subtree_visitor(-1,
    lambda kind, left, right:
        None if kind[0] is None else
        ["l", kind] + left + right + [(None, "*")]
)

# new visitor
c.append_r_class_right_parent_subtree_visitor(-1,
    lambda kind, parent_kind, left, right:
        None if kind[0] is not None else
        ["l", parent_kind] + [left[0] + ("*",)] + left[1:] + right + [None]
)

# new visitor
c.append_r_class_left_parent_subtree_visitor(-1,
    lambda kind, parent_kind, left, right:
        None if kind[0] is not None else
        ["l", parent_kind] + left + [right[0] + ("*",)] + right[1:] + [None]
)

# old visitor
c.append_r_class_right_parent_subtree_visitor(-1,
    lambda kind, parent_kind, left, right:
        None if kind[0] is not None else
        ["s", (None, "*"), parent_kind] + left + right
)

# new visitor
c.append_r_class_left_parent_subtree_visitor(-1,
    lambda kind, parent_kind, left, right:
        None if kind[0] is not None else
        ["s", None, parent_kind] + left + [right[0] + ("*",)] + right[1:]
)

# new visitor
c.append_r_class_right_parent_subtree_visitor(-1,
    lambda kind, parent_kind, left, right:
        None if kind[0] is not None else
        ["s", None, parent_kind] + [left[0] + ("*",)] + left[1:] + right
)
\end{lstlisting}

>>>>

\title{Combinatorial interpretations of~holonomic recurrences}
\author{Bartłomiej Puget}

@@

\DeclareMathOperator{\textiff}{\text{iff}}
\DeclareMathOperator{\N}{\mathbb{N}}
\DeclareMathOperator{\poly}{\mathbb{P}}
\DeclareMathOperator{\troot}{\text{root}}
\DeclareMathOperator{\tleft}{\text{left}}
\DeclareMathOperator{\tright}{\text{right}}
\DeclareMathOperator{\tLeft}{\text{Left}}
\DeclareMathOperator{\tRight}{\text{Right}}
\DeclareMathOperator{\tSub}{\text{Sub}}
\DeclareMathOperator{\tPre}{\text{Pre}}
\DeclareMathOperator{\tSuf}{\text{Suf}}
\DeclareMathOperator{\tClass}{\text{Class}}
\DeclareMathOperator{\tSection}{\text{Section}}
\DeclareMathOperator{\tPart}{\text{Part}}
\DeclareMathOperator{\n}{\bullet}
\DeclareMathOperator{\no}{\o}

@@

Generating functions is a way of describing infinite sequences using formal power series, where coefficients encode the consecutive elements. It allows us to incorporate many advanced methods of analysis of encoded objects.

Most recursive structure can be described using a~generating function equation, which can be then solved and transformed into a~holonomic recurrence, i.e.~recurrence that in order to define \(n\)-th element of the sequence requires just finite number of previous ones and coefficients that are polynomials in \(n\).

It allows us to quickly count objects of given structure with given size. At first glance, the recurrence seems to be closely related to the structure itself. It just seems like it should describe simple transformations resulting in constructing bigger objects from slightly smaller ones. In case of trees it could be for example adding nodes. However, there is no known method of interpreting it in the language of the combinatorial structures.

@@

The initial inspiration for the thesis came from the publication \textit{A natural counting of lambda terms}~\cite{inspiration}, which shows bijections among seemingly different tree structures, which are lambda trees, black-white trees~\cite{blackwhite} and zigzag free trees~\cite{inspiration}\todo{Find better cite}. They share the same holonomic recurrence, but there is no known immediate connection between the holonomic recurrence and the structures themselves.

Lambda terms are very popular structures used for analyzing functions and programming languages. Their very simple recursive definition makes them very attractive. A better understanding of them could allow us improving programming language compilers and in consequence improve every aspect of our lives. The interpretation of holonomic recurrence for lambda trees itself could allow us to discover new ways of generating random lambda terms via extending smaller ones in similar way to Rémy's algorithm for sampling binary trees~\cite{remy,note}.

@@

Section~\ref{sec:brutal_iterative_approach} presents our final approach to the original problem, i.e.~computer-aided iterative approach that allowed us to find the intermediate solution. We believe that it is possible to extend this idea in order to obtain the full solution, but we were not able to achieve this due to the limited time that was available to us.

@@

\(\troot(T)\) is implicitly defined as the~root of the~tree \(T\).

\(\tleft(v)\) is implicitly defined as the~left child of the~node \(v\).

\(\tright(v)\) is implicitly defined as the~right child of the~node \(v\).

\(\tSub(v)\) is implicitly defined as the~whole subtree rooted with the~node \(v\).

\(\tLeft(v)\) is implicitly defined as the~left subtree of the~node \(v\).

\(\tRight(v)\) is implicitly defined as the~right subtree of the~node \(v\).

Sibling of the node \(v\) is defined as the node \(u\) having the same parent as \(v\).

Uncle of the node \(v\) is defined as the~sibling of the parent of \(v\).

@@

We say that node \(v\) is a~twinling \(\textiff\) \(v\) has at least one sibling.

@@

\(x_i\) is implicitly assumed to be the~\(i\)-th element of a sequence \(X\), where \(x\) can be any lower-case letter and \(X\) is upper-case \(x\).

@@

\begin{definition}
    The ordinary generating function of a sequence \(A = {(a_i)}_{i=0}^{\infty}\) is defined as the following formal power series:
\[\gf{A}(z) = \sum_{i=0}^{\infty} a_i z^i.\]
\end{definition}

Note that in order to compute the \(k\)-th element of the sequence, namely \(a_k\), we just need to find the~constant corresponding to indeterminate \(z^k\).

@@

\begin{example}%
    \label{ex:bin_gf}
    Let's consider the recursive definition of binary trees shown in Figure~\ref{fig:binary_recursion}. We can create a~functional equation to count trees with exactly \(n\) internal nodes. To do so, we make internal nodes contribute to the generating function with \(z^1 = z\) and leaves with \(z^0 = 1\):

@@

In the thesis, if not stated otherwise, we will be working on rooted ordered unlabeled \(n\)-ary trees. To understand the terms, let's define them:
\begin{itemize}
    \item rooted tree is a~tree with single node marked as root. This allows us to define hierarchy of nodes using such terms as parent and child;
    \item ordered trees introduce order on children from the leftmost to the rightmost;
    \item unlabeled trees have nodes that are not distinguishable by their label. We will introduce soon kinds of nodes, but you shouldn't treat them as labels, but rather way to describe the structure of the node that is already present (e.g.~we can call node with just left child an~\(l\) node, but it is not its label. If the right child was added to the \(l\) node, it would not be an~\(l\) node anymore).
\end{itemize}
Additionally we allow that \(n\)-ary node has some of its children unset (e.g.~we can define a~binary node with just left child set, but not right one).

@@

\begin{definition}
    For single node \(v\), the pre-order notation is defined as \(\enc{v}\).
\end{definition}

@@

\begin{definition}
    Let \(n\) be the length of the sequence \(\enc{T}\). Notation of suffix of \(T\) in relation to \(v\), namely \(\enc{\tSuf(T, v)}\), is defined as a~suffix of \enc{T} of size \(n - \size{\enc{\tPre(T, v)}} - \size{\enc{\tSub(v)}}\). Note that the suffix may be an ordered forest.\todo{add to the example}
\end{definition}

@@

These interpretations have been source of ideas that we were incorporating in order to make our partial interpretation for the case of lambda trees in unary de Bruijn notation.

@@

Most of the time we are operating on trees, we define the size as a count of some objects (nodes or edges). We can assign weights to these objects and make the size a weighted sum. The weights can be variable or we can assign some other constant values (e.g.~small prime numbers).

@@

To show the idea of such transformations, let's define transformation context with holes.

\begin{definition}
    The transformation context is a rooted tree. Some of nodes can be left as holes, that may be later filled with trees. It also specifies type of the edge above the root of the context (i.e.~if that is actual edge, optional actual edge, virtual edge, or if there is no edge at all).
\end{definition}

\begin{example}%
    \label{ex:transformation}%
    We want to modify a node being right a child of its parent. We do not care what is below that node (i.e.~if that is a leaf or if it has children). We want to add a new node of which the original one is left child. We do not want the new node to have a right child. Additionally we want to swap potential children of the original node. We can define following transformation:
    \begin{center}
        \includeinlinescaledsvg{.4}{.7}{lambda__contexts__def_001}%
        \(\Rightarrow\)
        \includeinlinescaledsvg{.4}{.7}{lambda__contexts__def_002}%
    \end{center}
\end{example}

\(H_1\), \(H_2\), and \(H_3\) are holes. Edge crossing the box around the context defines type of edge the context is connect to the rest of the tree (i.e.~optional actual edge). We marked the node being described to clarify the picture.

@@

\begin{itemize}
    \item node is a~virtual leaf (each tree with \(n\) actual nodes has exactly \((n + 1)\) places where such contex apply):\\
        \includeinlinescaledsvg{1}{.7}{lambda__contexts__type_001}%
    \item node is an~actual node (tree with \(n\) actual nodes has excatly \(n\) places where such context apply):\\
        \includeinlinescaledsvg{1}{.7}{lambda__contexts__type_002}%
    \item node is a~root (each tree has exactly \(1\) place where such context apply):\\
        \includeinlinescaledsvg{1}{.7}{lambda__contexts__type_003}%
    \item node has its parent on the left, i.e.~is right child of its parent (the node can be specified to be actual or virtual):\\
        \includeinlinescaledsvg{.5}{.7}{lambda__contexts__type_004}%
        \includeinlinescaledsvg{.5}{.7}{lambda__contexts__type_004b}%
    \item node has its parent on the right, i.e.~is the left child of its parent (the node can be specified to be actual or virtual):\\
        \includeinlinescaledsvg{.5}{.7}{lambda__contexts__type_005}%
        \includeinlinescaledsvg{.5}{.7}{lambda__contexts__type_005b}%
\end{itemize}

@@

Note that for each transformation we can specify number and kinds of the nodes added to the tree. Let's take a look at the transformation from Example~\ref{ex:transformation}:
\begin{center}
    \includeinlinescaledsvg{.4}{.7}{lambda__contexts__def_001}%
    \(\Rightarrow\)
    \includeinlinescaledsvg{.4}{.7}{lambda__contexts__def_002_with_l}%
\end{center}
It adds single \(l\) node.

We will be operating on transformation classes, sections. They will start being empty and we will be adding tranformations to them.

\(\tSection_\tau\), where \(\tau\) is list constructed from \(\{a, l, o, s\}\) (you can treat is as a~set, but with potential repetitions), stores transformation extending the tree with nodes \(\tau\).

\(\tClass_i\), where \(i \in \N\) is defined as \(\bigcup_{\tau : |\tau| = i} \tSection_\tau\) (i.e.~set of all transformations adding exactly \(i\)~nodes to the tree).

We will be also talking about trees generated by the transformation. Let's consider constant \(n \in \N\). List of trees generated by the transformation from \(\tSection_\tau\) is defined as all trees generated by applying the transformation to all trees of size \((n - |\tau|)\).

Analogically, we can specify list of trees generated by the \(\tSection_\tau\) and \(\tClass_i\) as list of trees generated by all transformations from them.

Each section is associated with an expression coming from the holonomic recurrence. It defines expected number of trees that will be generated by each of them. Let \(k = n - |\tau|\) for each \(\tSection_\tau\). In the case of trees of lambda terms, we get following expressions associated to each section:

\[\begin{array}{lcr}
        \tSection_{\emptyset} &:& (k + 1) t_k\\
        \tSection_{\{l\}} &:& - (2 k + 1) t_k\\
        \tSection_{\{s\}} &:& - 2 (k + 1) t_k\\
        \tSection_{\{ao\}} &:& - (4 k) t_k\\
        \tSection_{\{ll\}} &:& (k) t_k\\
        \tSection_{\{ls\}} &:& (4 k + 2) t_k\\
        \tSection_{\{ss\}} &:& (k + 1) t_k\\
        \tSection_{\{aos\}} &:& 2 (2 k + 1) t_k\\
        \tSection_{\{lls\}} &:& - (2 k) t_k\\
        \tSection_{\{lss\}} &:& - (2 k + 1) t_k\\
        \tSection_{\{llss\}} &:& (k) t_k\\
\end{array}\]

Using \(k\) instead of \(n\) makes it easier to interpret, as we immediately see number of trees that should be generated by each section. For example \(\tSection_{\{ll\}}\) should generate \(k t_k\) trees from trees of size \(k\), which can be interpreted as taking each actual node and adding two \(l\) nodes above it.

Note that some of these expressions will evaluate to negative numbers. It is because of the fact that sum of all sections is equal to zero. We will be generating two kinds of trees then: positive (contributing to the total number of trees being generated as \(1\)) and negative ones (contributing as \(-1\)).

@@

As we want to make our work kind of organized, we are going to divide the process of finding interpretation into iterations. Each iteration should be a~small step towards solving the original problem.

Our brains like small examples, so we will consider small constant \(n = 3\). It will be describing size of the trees we want to generate. Because we start with such small \(n\), some of expressions associated with some sections will be empty, but it is ok. That's less trees to look at and soon we will increase \(n\).

We will be still using \(k = n - |\tau|\) when considering sections to make the view simpler.

Let's take a look at the number of trees we should generate:

\begin{lstlisting}
=== Classes stats ===
[ ]  0,0:       0 of      16 | (k + 1) * Tk              []
[ ]  1,0:       0 of     -10 | - (2 * k + 1) * Tk        [l]
[ ]  1,1:       0 of     -12 | - 2 * (k + 1) * Tk        [s]
[ ]  2,0:       0 of      -4 | - (4 * k) * Tk            [ao]
[ ]  2,1:       0 of       1 | (k) * Tk                  [ll]
[ ]  2,2:       0 of       6 | (4 * k + 2) * Tk          [ls]
[ ]  2,3:       0 of       2 | (k + 1) * Tk              [ss]
[ ]  3,0:       0 of       2 | 2 * (2 * k + 1) * Tk      [aos]
[x]  3,1:       0 of       0 | - (2 * k) * Tk            [lls]
[ ]  3,2:       0 of      -1 | - (2 * k + 1) * Tk        [lss]
[x]  4,0:       0 of       0 | k * Tk                    [llss]
\end{lstlisting}

The listing above is statistics view from Transformator framework. Let's analyze the first line:
\begin{itemize}
    \item \verb|[ ]| at the beginning means that we still need to generate some more trees. If we have generated all of them, it would be \verb|[x]|;
    \item \verb|0,0| is class and section identifier, i.e. \(|\tau|\) and arbitrarily assigned section identifier;
    \item \verb|0 of 16| means that we have generated \(0\) trees from \(16\) expected;
    \item \verb|(k + 1) * Tk| is just the expression associated with the section;
    \item \verb|[]| at the end is \(\tau\), which in this case is equal to \(\emptyset\).
\end{itemize}

We start with interpreting \(\tSection_\emptyset\). From statistics above, we see that we should generate \(16\) trees. \((k + 1)\) suggests pointing to a~virtual leaf. Let's do so. We do that by defining following transformation:

\begin{center}
    \includeinlinescaledsvg{.4}{.7}{lambda__transformations__001a}%
    \(\xRightarrow{+1}\)%
    \includeinlinescaledsvg{.4}{.7}{lambda__transformations__001b}%
\end{center}

Here \(\xRightarrow{+1}\) means that we are generating positive trees.

In the language of Transformator framework it would be:

@@

After such operation we get following statistics:

\begin{lstlisting}
# define 0,0: (k + 1) * Tk []
=== Classes stats ===
[x]  0,0:      16 of      16 | (k + 1) * Tk              []
[ ]  1,0:       0 of     -10 | - (2 * k + 1) * Tk        [l]
[ ]  1,1:       0 of     -12 | - 2 * (k + 1) * Tk        [s]
[ ]  2,0:       0 of      -4 | - (4 * k) * Tk            [ao]
[ ]  2,1:       0 of       1 | (k) * Tk                  [ll]
[ ]  2,2:       0 of       6 | (4 * k + 2) * Tk          [ls]
[ ]  2,3:       0 of       2 | (k + 1) * Tk              [ss]
[ ]  3,0:       0 of       2 | 2 * (2 * k + 1) * Tk      [aos]
[x]  3,1:       0 of       0 | - (2 * k) * Tk            [lls]
[ ]  3,2:       0 of      -1 | - (2 * k + 1) * Tk        [lss]
[x]  4,0:       0 of       0 | k * Tk                    [llss]
=== Diff stats ===
0,0:     0    16
\end{lstlisting}

Note that we have generated all expected trees associated with \(\tSection_\emptyset\) (it is indicated by \verb|[x]|).

We also get the \verb|Diff stats|. It shows us information about how we have unbalanced the positive and negative side of our equation. At the end we want to generate the same list of trees at the negative and positive site.

The framework also can provide us with the visualization of the trees we have generated. For example, from the tree on the left side of \(\xRightarrow{+1}\) we have generated four trees:

\includeinlinesvg{.19}{lambda__trees_00__4_base}%
\(\xRightarrow{+1}\)%
\includeinlinesvg{.19}{lambda__trees_00__4}%
\includeinlinesvg{.19}{lambda__trees_00__5}%
\includeinlinesvg{.19}{lambda__trees_00__6}%
\includeinlinesvg{.19}{lambda__trees_00__7}%

You can see that there are four virtual leaves matching left side of the transformation we have defined.

We can also query Transformator framework to visualize the trees that are unbalanced. For \(\tClass_0\) those are:

@@

We have ordered the trees and enumerated the lines to make the analysis clearer.

We need to generate all of these trees on the negative side of the equation. Let's take a look at the line (1). All of these trees share the same property, that the virtual leaf being pointed is child of \(l\)~node. Analogically at the line (2) pointed virtual leaves are children of \(s\) nodes. To generate such trees on the negative side we will use \(\tSection_{\{l\}}\) and \(\tSection_{\{s\}}\).

The transformation to resolve case of the line (1) is simple:

\begin{center}
    \includeinlinescaledsvg{.4}{.7}{lambda__transformations__002a}%
    \(\xRightarrow{-1}\)%
    \includeinlinescaledsvg{.4}{.7}{lambda__transformations__002b}%
\end{center}

\begin{lstlisting}
c.append_r_class_subtree_visitor(-1,                # negative node
    lambda kind, left, right:
        None if kind[0] is None else                # we ignore all virtual nodes
        ["l", kind] + left + right + [(None, "*")]  # for each actual node we add
                                                    # l node with right virtual
                                                    # node pointed as a parent
)
\end{lstlisting}

We just add an \(l\) node above each actual one and point its right child being virtual leaf.

It creates following trees:

\includeinlinesvg{.24}{lambda__trees_100__0_base}\(\xRightarrow{-1}\)\includeinlinesvg{.24}{lambda__trees_100__0}%
\includeinlinesvg{.24}{lambda__trees_100__1_base}\(\xRightarrow{-1}\)\includeinlinesvg{.24}{lambda__trees_100__1}%

\includeinlinesvg{.24}{lambda__trees_100__2_base}\(\xRightarrow{-1}\)\includeinlinesvg{.24}{lambda__trees_100__2}%
\includeinlinesvg{.24}{lambda__trees_100__3_base}\(\xRightarrow{-1}\)\includeinlinesvg{.24}{lambda__trees_100__3}%

It addresses all cases from the line (1), but also introduces an~incorrect lambda tree in which \(s\) node is the parent of \(l\) node, which cannot happen (i.e.~such tree does not correspond to valid lambda term). To mitigate this problem we will later use \(\tSection_{\{ls\}}\) and generate positive trees where \(s\) node is parent of \(l\) node.

To generate negative trees corresponding to the line (2), create following transformation for \(\tSection_{\{s\}}\).

\begin{center}
    \includeinlinescaledsvg{.4}{.7}{lambda__transformations__003a}%
    \(\xRightarrow{-1}\)%
    \includeinlinescaledsvg{.4}{.7}{lambda__transformations__003b}%
\end{center}

\begin{lstlisting}
c.append_r_class_right_parent_subtree_visitor(-1,
    lambda kind, parent_kind, left, right:
        None if kind[0] is not None else
        ["s", (None, "*"), parent_kind] + left + right
)
\end{lstlisting}

It generates exactly the trees we need:

\begin{center}
    \includeinlinescaledsvg{.24}{.4}{lambda__trees_110__0_base}%
    \(\xRightarrow{-1}\)%
    \includeinlinescaledsvg{.24}{.4}{lambda__trees_110__0}%

    \includeinlinescaledsvg{.24}{.4}{lambda__trees_110__1_base}%
    \(\xRightarrow{-1}\)%
    \includeinlinescaledsvg{.24}{.4}{lambda__trees_110__1}%
    \includeinlinescaledsvg{.24}{.4}{lambda__trees_110__2_base}%
    \(\xRightarrow{-1}\)%
    \includeinlinescaledsvg{.24}{.4}{lambda__trees_110__2}%
\end{center}

Next, we can take care of cases from line (3). To do so, we will use both \(\tSection_{\{l\}}\) and \(\tSection_{\{s\}}\).

\begin{center}
    \includeinlinescaledsvg{.4}{.7}{lambda__transformations__004a}%
    \(\xRightarrow{-1}\)%
    \includeinlinescaledsvg{.4}{.7}{lambda__transformations__004b}%
\end{center}
